$\pi$ with leftovers: \\ a mechanisation in Agda

### 2021-04-13 paper/paper.tex
Add references to work on usage algebras

\begin{abstract}
  Linear type systems need to keep track of how programs use their resources.
  The standard approach is to use \emph{context splits} specifying how resources are (disjointly) split across subterms.
  In this approach, context splits redundantly echo information which is already present within subterms.  An alternative approach is to use \emph{leftover typing} \cite{Mackie,Allais2018a}, where in addition to the usual (input) usage context, typing judgments have also an output usage context: the leftovers.
  In this approach, the leftovers of one typing derivation are fed as input to the next, threading through linear resources while avoiding context splits.
  We use leftover typing to define a type system for a {resource-aware \picalc{}} \cite{MilnerPW92,Milner99}, a process algebra used to  model concurrent systems.
  Our type system is parametrised over a set of \emph{usage algebras} that are general enough to encompass \emph{shared types} (free to reuse and discard), \emph{graded types} (use exactly $n$ number of times) and \emph{linear types} (use exactly once).
   Linear types are important in the \picalc{}: they ensure privacy and safety of communication and avoid race conditions, while graded and shared types allow for more flexible programming.
  We provide a framing theorem for our type system, generalise the weakening and strengthening theorems to include linear types, and prove subject reduction.
  Our formalisation is fully mechanised in about 1850 lines of Agda \cite{Zalakain2020Agda}.

@@

\paragraph*{Contributions and Structure of the Paper}
\begin{enumerate}
  \item \textbf{Leftover typing for resource-aware \picalc{}}. Our type system uses leftover typing to model the resource-aware \picalc{} (\autoref{leftover-typing}) and satisfies subject reduction (\autoref{thm:subject-reduction}).
    In addition to making context splits unnecessary, leftover typing allows for a \emph{framing} theorem (\autoref{thm:framing}) to be stated and is naturally associative, making type safety properties considerably easier to reason about (\autoref{meta-theory}).
    Thanks to leftover typing, we can now state \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the whole framework, not just the shared fragment. This give a uniform and complete presentation of all the meta-theory for the resource-aware \picalc{}.    
  \item \textbf{Shared, graded and linear unified \picalc{}}.
  We {generalise resource counting to a set of usage algebras that can be mixed within the same type system}.
    We do not instantiate our type system to only work with linear resources, instead we present an algebra-agnostic type system, and admit a mix of user-defined \emph{resource aware} algebras (\autoref{multiplicities}).
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra.
    Multiple algebras can be mixed in the type system --- usage contexts keep information about what algebra to use for each type (\autoref{contexts}).
    In particular, this allows for type systems combining linear (use exactly once), graded (exact number of $n$ times) and shared (free to reuse and discard) types under the same framework.

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
We generalise over this notion by defining an algebra for multiplicities that is satisfied by linear, graded and shared types alike.
We then use pairs of multiplicities as usage annotations for a channel's input and output capabilities.

@@

Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq, where de Bruijn indices are less popular than in Agda or Idris.
They use their tool to help automate proofs of subject reduction for a type system with session types.
%
Orchard and Yoshida \cite{OrchardY16} embed a small effectul imperative language into the session-typed \picalc{}, showing that session types are expressive enough to encode effect systems.
%
Based on contextual type theory \cite{Pientkaa,Pientka}, LINCX \cite{Georges2017} extends the linear logical framework LLF \cite{Cervesato1996} by internalising the notion of bindings and contexts.
The result is a meta-theory in which HOAS encodings with both linear and dependent types can be described.
The developer obtains for free an equational theory of substitution and decidable typechecking without having to encode context splits within the object language.
%
Further work on mechanisation of the \picalc{} \cite{Henry-Gerard1999,Honsell2001a,Bengtson2013,Despeyroux2000,Affeldt2008}, focuses on non-linear variations, differently from our range of linear, graded and shared types.

>>>>

\begin{abstract}
  Linear type systems need to keep track of how programs use their resources.
  The standard approach is to use \emph{context splits} specifying how resources are (disjointly) split across subterms.
  In this approach, context splits redundantly echo information which is already present within subterms.  An alternative approach is to use \emph{leftover typing} \cite{Mackie,Allais2018a}, where in addition to the usual (input) usage context, typing judgments have also an output usage context: the leftovers.
  In this approach, the leftovers of one typing derivation are fed as input to the next, threading through linear resources while avoiding context splits.
  We use leftover typing to define a type system for a {resource-aware \picalc{}} \cite{MilnerPW92,Milner99}, a process algebra used to  model concurrent systems.
  Our type system is parametrised over a set of \emph{usage algebras} \cite{JungSSSTBD15,TuronTABD13} that are general enough to encompass \emph{shared types} (free to reuse and discard), \emph{graded types} (use exactly $n$ number of times) and \emph{linear types} (use exactly once).
   Linear types are important in the \picalc{}: they ensure privacy and safety of communication and avoid race conditions, while graded and shared types allow for more flexible programming.
  We provide a framing theorem for our type system, generalise the weakening and strengthening theorems to include linear types, and prove subject reduction.
  Our formalisation is fully mechanised in about 1850 lines of Agda \cite{Zalakain2020Agda}.

@@

\paragraph*{Contributions and Structure of the Paper}
\begin{enumerate}
  \item \textbf{Leftover typing for resource-aware \picalc{}}. Our type system uses leftover typing to model the resource-aware \picalc{} (\autoref{leftover-typing}) and satisfies subject reduction (\autoref{thm:subject-reduction}).
    In addition to making context splits unnecessary, leftover typing allows for a \emph{framing} theorem (\autoref{thm:framing}) to be stated and is naturally associative, making type safety properties considerably easier to reason about (\autoref{meta-theory}).
    Thanks to leftover typing, we can now state \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the whole framework, not just the shared fragment. This give a uniform and complete presentation of all the meta-theory for the resource-aware \picalc{}.    
  \item \textbf{Shared, graded and linear unified \picalc{}}.
  We {generalise resource counting to a set of usage algebras that can be mixed within the same type system}.
    We do not instantiate our type system to only work with linear resources, instead we present an algebra-agnostic type system, and admit a mix of user-defined \emph{resource aware} algebras \cite{JungSSSTBD15,TuronTABD13} (\autoref{multiplicities}).
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra.
    Multiple algebras can be mixed in the type system --- usage contexts keep information about what algebra to use for each type (\autoref{contexts}).
    In particular, this allows for type systems combining linear (use exactly once), graded (exact number of $n$ times) and shared (free to reuse and discard) types under the same framework.

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 (exhausted) or 1 (available).
We generalise over this notion by defining an algebra for multiplicities \cite{JungSSSTBD15,TuronTABD13} that is satisfied by linear, graded and shared types alike.
We then use pairs of multiplicities as usage annotations for a channel's input and output capabilities.

@@

Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq, where de Bruijn indices are less popular than in Agda or Idris.
They use their tool to help automate proofs of subject reduction for a type system with session types.
%
Orchard and Yoshida \cite{OrchardY16} embed a small effectul imperative language into the session-typed \picalc{}, showing that session types are expressive enough to encode effect systems.
%
Based on contextual type theory, LINCX \cite{Georges2017} extends the linear logical framework LLF \cite{Cervesato1996} by internalising the notion of bindings and contexts.
The result is a meta-theory in which HOAS encodings with both linear and dependent types can be described.
The developer obtains for free an equational theory of substitution and decidable typechecking without having to encode context splits within the object language.
%
Further work on mechanisation of the \picalc{} \cite{Henry-Gerard1999,Honsell2001a,Bengtson2013,Despeyroux2000,Affeldt2008}, focuses on non-linear variations, differently from our range of linear, graded and shared types.
